Haskell in Computational Biology

### 2013-05-14 jfp.tex
line edits for jfp draft
closes #12

@@

\newcommand\figref[1]{Figure~\ref{#1}}
\newcommand\secref[1]{Section~\ref{sec:#1}}
\newcommand\seclabel[1]{\label{sec:#1}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell: 
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\looseness=-1
\end{itemize}

@@

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
\else
\centerline{\includegraphics[width=6cm]{alignment.eps}} 
\fi

\caption{A~structural alignment of four proteins $(\alignwidth = 12)$}
\label{alignment} 
\end{figure}

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid by deletion from column~$j$.
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.

@@

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
\ifpagetuning
``Plan7'' implies that
\else
The model gets its name because
\fi
 there are exactly~7 possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

@@

\label{plan7} \end{figure}

\begin{figure*}
\newcommand\vsum[2]{#2&{}+{}& #1}

\def\goo{18pt}
\def\gum{14pt}
\[
\def\maxiquad{\hskip 1.2em\relax}
%\mskip -5mu
\begin{array}{@{}l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
  \end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
  \vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
  \end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
  \vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
  \end{array} \right.\\
\end{array}
%
\hskip 1.5em\relax % \qquad too big
%
\begin{array}{l@{}c@{}l@{}}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
  \end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
  \vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
  \end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
  \vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
  \end{array} \right.\\[\gum]
%
\multicolumn3{l@{}}{%
  a'_{s \hat{s}} = - \log a_{s \hat{s}} 
\qquad
  e^{\prime}_{s}(x) = - \log\frac{e_{s}(x)}{q_{x}}
\qquad
  V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
%\mskip -5mu
\]

\caption{Viterbi's equations, in original and negated forms}
\label{viterbi}
\end{figure*}

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
\ifnotcutting and evolutionary history \fi
with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
The~path that maximizes the
log of the probability is the most likely path.

The computation is specified on the left-hand side of \figref{viterbi}.
A~probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Probabilities $V_j^I(i)$ and $V_j^D(i)$ are similar.
The
equations are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

@@

In a real hidden Markov model, each probability is represented as a score.
Our code implements a transformed version of Viterbi's equations
which operates on scores.
The transformed equations are
shown on the right-hand side of \figref{viterbi}.
They minimize the
score (the negated log probability) for each combination of
column~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or~$D_j$.

A~model is represented as a sequence of \emph{nodes};
node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
as well as the probabilities of transitions out of that node.
Each node contains tables of emission scores
$e'_{M_j}$~and~$e'_{I_j}$.
These tables are read by function
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
%
We place the transition probabilities 
into a record in which each field is labeled
\texttt{$s$\_$\mskip 1mu\hat s$},
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels: 
\smallverbatiminput{tprob-tprobs}
These scores are read by
function \texttt{aScore}, 
whose specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we define
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The~sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
\smallfuzzverbatiminput{2.9pt}{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown in \figref{viterbi}:
\mrfy\ computes not only the {probability} of the most likely path but 
also the path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+.
Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
but faster: 
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

Memoization makes \verb+vee'+ perform as well as our classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to dynamic programming.
By~contrast, standard implementations of Viterbi's algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic version.

@@

We use Viterbi's algorithm to help detect
homologies in proteins with specific kinds of structure.
When a real protein folds in three dimensions, 
amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
Some groups of such acids are called \emph{beta strands}.
Beta strands
can be hydrogen-bonded to each other,
making them ``stuck together.''
These beta strands help identify groups of homologous
proteins.
MRFy detects homologous proteins that include hydrogen-bonded beta
strands; using prior methods, many instances of this problem are
intractable.

@@

@@

%
%  placement at top left column is not ideal, but right column results
%  in a bad page break elsewhere
%
\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.eps}} 
\fi
Each shaded node represents a beta-strand position.
Nodes connected by dashed edges are hydrogen-bonded.

\caption{A Markov random field with two beta-strand pairs}
\label{mrf} 
\end{figure}

@@

@@

\begin{figure}%%%[h!]
\centerline{\includegraphics{speedup}}
\medskip
\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
\label{fig:speedup}
\end{figure}

>>>>

\ifpdfmadness
  \def\loadpdf[#1]#2{\includegraphics[#1]{#2.pdf}}
\else
  \def\loadpdf[#1]#2{\includegraphics[#1]{#2.eps}}
\fi

@@

\newcommand\figref[1]{Figure~\ref{fig:#1}}
\newcommand\figrefpair[3][and]{Figures \ref{fig:#2}~#1~\ref{fig:#3}}
\newcommand\figlabel[1]{\label{fig:#1}}
\newcommand\secref[1]{Section~\ref{sec:#1}}
\newcommand\seclabel[1]{\label{sec:#1}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\item
\emph{[POTENTIAL BULLETS ABOUT SPEED AND ``MORE FUNCTIONAL'']}
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds to the original paper by describing our experience in
trying to simplify and speed up the central algorithm.

@@

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=8cm]{alignment.pdf}} 
\else
\centerline{\includegraphics[width=8cm]{alignment.eps}} 
\fi

\caption{A~structural alignment of four proteins $(\alignwidth = 12)$}
\figlabel{alignment} 
\end{figure}

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid by deletion from column~$j$.
\hfuzz=1.5pt % fix blot
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.

@@

@@

\figlabel{plan7}
\end{figure}

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
\seclabel{plan7}%
\remark{Reference to \figref{plan7}??}%
\ifpagetuning
``Plan7'' implies that
\else
The model gets its name because
\fi
 there are exactly~7 possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
\ifnotcutting and evolutionary history \fi
with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by~summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
The~path that maximizes the
log of the probability is the most likely path.

\def\goo{18pt}
\def\gum{14pt}
\newcommand\vsum[2]{#2&{}+{}& #1}

\begin{figure}

\[
\def\maxiquad{\hskip 1.2em\relax}
%\mskip -5mu
\begin{array}{@{}l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
  \end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
  \vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
  \end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
  \vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
  \end{array} \right.\\
\end{array}
\]

\caption{Viterbi's original equations}
\figlabel{viterbi-original}

\bigskip

%\begin{figure}
\[
\begin{array}{l@{}c@{}l@{}}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
  \end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
  \vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
  \end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
  \vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
  \end{array} \right.\\[\gum]
%
\multicolumn3{l@{}}{%
  a'_{s \hat{s}} = - \log a_{s \hat{s}} 
\qquad
  e^{\prime}_{s}(x) = - \log\frac{e_{s}(x)}{q_{x}}
\qquad
  V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

\caption{Viterbi's equations transformed to work with scores}
\figlabel{viterbi-transformed}
\end{figure}

The computation is specified in \figref{viterbi-original}.
A~probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Probabilities $V_j^I(i)$ and $V_j^D(i)$ are similar.
The
equations are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

@@

In a real hidden Markov model, each probability is represented as a score.
Our code implements a transformed version of Viterbi's equations
which operates on scores.
The~transformed equations are
shown in \figref{viterbi-transformed}.
They minimize the
score (the negated log probability) for each combination of
column~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or~$D_j$.

A~model is represented as a sequence of \emph{nodes};
node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
as well as the probabilities of transitions out of that node.
Each node contains tables of emission scores
$e'_{M_j}$~and~$e'_{I_j}$.
These tables are read by function
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
%
We place the transition probabilities 
into a record in which each field is labeled
\texttt{$s$\_$\mskip 1mu\hat s$},
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels: 
\smallverbatiminput{tprob-tprobs}
The fields of this record are read by
function \texttt{aScore}, 
which is given states $s$~and~$\hat s$ and a node index; its %% fixes a bad line break
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
of \figref{viterbi-transformed}, we define
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The~sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
\smallfuzzverbatiminput{2.9pt}{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown
in \figrefpair[or]{viterbi-original}{viterbi-transformed} :
\mrfy\ computes not only the {probability} of the most likely path but 
also the path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+.
Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
but faster: 
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
package
\path{Data.MemoCombinators}.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

Memoization makes \verb+vee'+ perform as well as our classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to dynamic programming.
By~contrast, standard implementations of Viterbi's algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi-transformed}, we~retired the classic version.

@@

We use Viterbi's algorithm to help detect
homologies in proteins with specific kinds of structure.
When a real protein folds in three dimensions, 
amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
Some groups of such acids are called \emph{beta strands}.
Beta strands
can be hydrogen-bonded to each other,
making them ``stuck together.''
These beta strands help identify groups of homologous
proteins.
MRFy's \emph{raison d'\^etre} is to detect homologous proteins that
include hydrogen-bonded beta 
strands; using prior methods, many instances of this problem are
intractable \cite{daniels:thesis}.

@@

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=11cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=11cm]{mrf_interleave_diagram.eps}} 
\fi
Each shaded node represents a beta-strand position.\\
Nodes connected by dashed edges are hydrogen-bonded.

\smallskip

\caption{A Markov random field with two beta-strand pairs}
\figlabel{mrf} 
\end{figure}

@@

@@

\begin{figure}
\tabcolsep=0.5cm
\centerline{%
\begin{tabular}{cc}
\loadpdf[width=5cm]{search-figs/search-graph1}&
\loadpdf[width=5cm]{search-figs/search-graph5}\\
\begin{minipage}{4cm}
\caption{Stochastic search with useful and useless steps}%
\figlabel{search-steps}
\end{minipage}%
&
\begin{minipage}{4cm}
\caption{Filtering out useless steps could cause nontermination}%
\figlabel{filtered-search-steps}
\end{minipage}\\
\end{tabular}%
}
\end{figure}

@@

\begin{figure}%%%[h!]
\centerline{\includegraphics{speedup}}
\medskip
\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
\figlabel{speedup}
\end{figure}
